\modLP{Identification of} Biological Regulatory Networks from Process Hitting models

### 2014-11-12 response.tex

\begin{document}

\section*{Summary of the main changes}

We thank the reviewer for these helpful comments and suggestions.
The reviewer's main concerns about our work were the readability of our paper and the consistency problems posed by our definition of Thomas' modeling.

We made the following main changes to the article:
\begin{itemize}
  \item The title has been changed from “Constructing...” to “Identification of...”.
  \item The abstract and the introduction have been improved to better contextualize our work.
  \item Additional explanations have been added at the beginning of every section
    to detail their objectives and methods.
  \item The explanations about cooperations in Process Hitting have been partially rewritten.
  \item In the section defining Thomas modeling (subsection 2.2):
    \begin{itemize}
      \item “Unsigned edges” are now called “non-monotonous edges”,
        and explanations have been added to explain their meaning;
      \item We now only consider single-value parameters
        (as opposed to previous interval parameters);
      \item A property has been added to characterize equivalent parametrizations
        (see Remark 1).
    \end{itemize}
  \item A new section has been created (section 3)
    in order to better separate general definitions 
    from the Interaction Graph (IG) inference propositions of section 4.
  %\item The IG inference method has been improved in order to get more accurate results
  %  (see Proposition 1).
  %\item Property 4 (Well-formed PH for parameter inference) has been improved
  %  to handle more general forms of PH models,
  %  namely, with cooperative sorts containing more than 4 processes.
  %\item Equations of section 5 (19 to 22) have been adapted to single-value parameters.
  \item Some definitions and propositions have been updated to make them more simple and readable
  %All examples in sections 2 to 5 have been actualized to the new formalism
  %  and propositions.
  \item A detailed example has been added in section 7 (see subsection 7.1)
    where we detail the application of our method to some parts of a model of
    the bacteriophage lambda immunity control.
  %\item Actualized results are given in subsection 7.3,
  %  related to several improvements brought to our method.
\end{itemize}

@@

%\TODO{Answer inc. text modification}

\question{
(P.1) Abstract: the following sentence is not understandable. Either more explanations
are needed or simply suppress it.
``Its major originality lies in a specific restriction on the causality of actions, which makes the formal
analysis of very large systems tractable.''
}

\answer
% \DISCUSS{LP}
% proposition: ``The PH is a particular class of Indeterministic Asynchronous Automata
% Network (or safe Petri nets) on which have been designed static analysis for
% dynamical properties that are tractable on very large systems.''

This comment has been taken into account and the sentence has been replaced by:

“The analysis of the dynamics of very large models is however a rather hard problem, which led us to previously introduce the Process Hitting framework (PH), which is a particular class of nondeterministic asynchronous automata network (or safe Petri nets). Its major advantage lies in the efficiency of several static analyses recently designed to assess dynamical properties, making it possible to tackle very large models.”

### 2014-11-12 main.tex
complete phage lambda example
todo--
fix \irB
phage lambda++
cleanup bib, make Fig 1 more visible, reference to K equivalence
plus de couleurs
minor corrections
last minor corrections

\usepackage{prettyref}
\newrefformat{def}{Def.~\ref{#1}}
\newrefformat{fig}{Fig.~\ref{#1}}
\newrefformat{pro}{Property~\ref{#1}}
\newrefformat{pps}{Proposition~\ref{#1}}
\newrefformat{lem}{Lemma~\ref{#1}}
\newrefformat{thm}{Theorem~\ref{#1}}
\newrefformat{sec}{Sect.~\ref{#1}}
\newrefformat{ssec}{Subsect.~\ref{#1}}
\newrefformat{sssec}{Subsect.~\ref{#1}}
\newrefformat{suppl}{Appendix~\ref{#1}}
\newrefformat{eq}{Eq.~\eqref{#1}}
\newrefformat{ex}{Example~\ref{#1}}
\newrefformat{tb}{Table~\ref{#1}}
\def\pref{\prettyref}

\newdefinition{definition}{Definition}
\newdefinition{property}{Property}
\newdefinition{proposition}{Proposition}
\newdefinition{example}{Example}

@@

\def\modMF#1{\textcolor{teal}{#1}}
\def\modLP#1{\textcolor{magenta}{#1}}
\def\modMM#1{\textcolor{blue}{#1}}
\def\modOR#1{\textcolor{olive}{#1}}

% Id est
\newcommand{\ie}{i.e.~}

@@

\begin{abstract}
\modMF{Maxime}
\modLP{Lo\"ic}
\modMM{Morgan}
\modOR{Olivier}

%\parindent 0.5cm
The Process Hitting (PH) \modLP{is a particular class of Indeterministic Asynchronous Automata
Network (or safe Petri nets) on which have been recently designed several static analyses for
dynamical properties that are tractable on very large systems.}

\modLP{In this paper, we address the formal identification of qualitative
models of Biological Regulatory Networks (BRNs) from PH models.}
First, the inference of the Interaction Graph (IG) from a PH model
\modLP{summarizes the signed influences between the components} that are effective for the dynamics.
Second, we provide the inference of all Ren\'e Thomas models of BRNs 
that are compatible with a given PH.
As the PH allows the specification of indeterministic interactions between
components, our inference emphasizes the ability of PH to deal with large BRNs with incomplete knowledge
on interactions, where Thomas' approach fails because of the combinatorics of parameters.

The inference of corresponding Thomas models is provided using Answer Set Programming,
which allows notably an efficient enumeration of (possibly numerous) compatible parametrizations.

%\todo{Rather name Kaufman as the creator of BRNs.}
%On the other hand, the qualitative modeling of BRNs has been widely addressed using Ren\'e Thomas'
%formalism, leading to numerous theoretical work and practical tools to understand emerging behaviors.
%In addition to be able to use the existing results on this formalism,
%obtaining the Thomas representation of a PH model allows to iteratively refine a model.
%Linking this representation to PH would not only allow to use the existing results on this formalism,
%but also to iteratively refine the model of a system.

%This paper establishes formal links between these two formalisms, permitting to appreciate their
%complementary features.
\end{abstract}

@@

\input{parts-infer-ig}
\input{parts-infer-param}
\input{parts-asp}
\input{parts-example}
\input{parts-discussion}

@@

%\input{parts/contributions}

>>>>

\usepackage{prettyref}
\newrefformat{def}{Def.~\ref{#1}}
\newrefformat{fig}{Fig.~\ref{#1}}
\newrefformat{pro}{Property~\ref{#1}}
\newrefformat{pps}{Proposition~\ref{#1}}
\newrefformat{lem}{Lemma~\ref{#1}}
\newrefformat{thm}{Theorem~\ref{#1}}
\newrefformat{sec}{Sect.~\ref{#1}}
\newrefformat{ssec}{Subsect.~\ref{#1}}
\newrefformat{sssec}{Subsect.~\ref{#1}}
\newrefformat{suppl}{Appendix~\ref{#1}}
\newrefformat{eq}{Eq.~\eqref{#1}}
\newrefformat{ex}{Example~\ref{#1}}
\newrefformat{tb}{Table~\ref{#1}}
\newrefformat{rem}{Remark~\ref{#1}}
\def\pref{\prettyref}

\newdefinition{definition}{Definition}
\newdefinition{property}{Property}
\newdefinition{remark}{Remark}
\newdefinition{proposition}{Proposition}
\newdefinition{example}{Example}

@@

\def\modMF#1{\textcolor{teal}{#1}}
\def\modLP#1{\textcolor{magenta}{#1}}
\def\modMM#1{\textcolor{blue}{#1}}
\def\modOR#1{\textcolor{olive}{#1}}
\def\modMF#1{#1} \def\modLP#1{#1} \def\modMM#1{#1} \def\modOR#1{#1}

% Abbréviations
\newcommand{\ie}{i.e.,\ }
\newcommand{\eg}{e.g.,\ }

@@

\begin{abstract}
\modOR{Qualitative formalisms offer a well-established alternative to the more traditionally used differential equation models of Biological Regulatory Networks (BRNs). These formalisms led to numerous theoretical works and practical tools to understand emerging behaviors.} \modMM{The analysis of the dynamics of very large models is however a rather hard problem, which led us to previously introduce} \modMF{the Process Hitting framework (PH),} \modMM{which} \modLP{is a particular class of nondeterministic asynchronous automata
network (or safe Petri nets).} \modMM{Its major advantage lies in the efficiency of several static analyses recently designed to assess dynamical properties, making it possible to tackle very large models.}

\modLP{In this paper, we address the formal identification of qualitative
models of BRNs from PH models.}
First, the inference of the Interaction Graph from a PH model
\modLP{summarizes the signed influences between the components} that are effective for the dynamics.
Second, we provide the inference of all Ren\'e Thomas models of BRNs 
that are compatible with a given PH.
As the PH allows the specification of nondeterministic interactions between
components, our inference emphasizes the ability of PH to deal with large BRNs with incomplete knowledge
on interactions, where Thomas' approach fails because of the combinatorics of parameters.

The inference of corresponding Thomas models is \modLP{implemented} using Answer Set Programming,
which allows \modMF{in particular} an efficient enumeration of (possibly numerous) compatible parametrizations.
\end{abstract}

@@

\input{parts-tr2global}
\input{parts-infer-ig}
\input{parts-infer-param}
\input{parts-asp}
\input{parts-example}
\input{parts-discussion}

@@

\appendix
\input{annex-phage-lambda}

### 2014-11-12 parts-asp.tex
last minor corrections

Cardinality rules turn out to be convenient to enumerate all possible parametrizations by creating multiple answer sets.
For functional purposes, a unique label is assigned to every possible set of resources of a given component.
Thus, we denote $\omega_p$ the set of resources of a given component $a$ labeled by $p$,
and naturally, $K_{a,\omega_p}$ is the related parameter.
%and in the following we note $K_{a,\omega_p}$ the parameter of component $a$ whose set of resources $\omega$ is assigned to the label $p$.
We note that labeling the sets of resources of a component is obviously equivalent to labeling its parameters.
Then, suppose that:
\begin{itemize}
  \item $\paramlabel(a, p)$ states that $p$ is a valid label for a set of resources of component $a$ (and therefore $K_{a,\omega_p}$ is a valid parameter);
  \item $\param(a, p, i)$ states that: $K_{a, \omega_p} = i$;
  \item $\inferedparam(a, p)$ states that the parameter inference of $K_{a, \omega_p}$ was conclusive (\pref{pps:param_K}).
\end{itemize}
It is thereby possible to enumerate the possible values of all parameters for which \pref{pps:param_K} was not conclusive, with the following cardinality rule:
\begin{align*}
  & 1\ \{\ \param(\var{A}, \var{P}, \var{I}) \mid \componentlevels(\var{A}, \var{I})\ \}\ \modMF{1}\ \la \\
  & \qquad\qquad\qquad \paramlabel(\var{A}, \var{P}), \nota \inferedparam(\var{A}, \var{P}).
\end{align*}
Indeed, this rule applies to any possible parameter $\var{P}$ of any component $\var{A}$ ($\paramlabel$) whose value is still unknown ($\nota \inferedparam$),
and states that any expression level $\var{I}$ of this component ($\componentlevels$)
is a candidate value for the parameter ($\param$).
\modMF{
Furthermore, both the lower and upper bounds are $1$,
which forces each enumerated parameter to have exactly one value.
}
%but no upper bound is specified ($\infty$) for the size of each parameter
%(which is in fact already bounded by the number of possible expression levels of the related component).
%each parameter contains at least $1$ value, as stated by the lower bound, and has no upper bound ($\infty$) but the number of expression levels of its component.
In other worlds, this cardinality rule creates as many answer sets as there are \emph{candidate} parametrizations
so that if $K_{a, \omega_p}$ could not be inferred by \pref{pps:param_K}, then
\modMF{
$K_{a, \omega_p} \in \segm{0}{l_a}$
}
(thus completely disregarding the notion of admissible parametrizations given in \pref{ssec:admissible-K}).

\begin{example}\label{ex:cardinality}
\modMF{%
In the scope of \pref{ex:infer-param-runningPH-1},
$K_{a,\{b\}}$ and $K_{a,\{c\}}$
could not be inferred by \pref{pps:param_K}.
The previous cardinality rule allows to produce 9 parametrizations,
in which these three parameters can take all possible values:
%(as, in this case, the assumptions of \pref{ssec:admissible-K} bring no new constraint):
\begin{align*}
  (K_{a,\{b\}}; K_{a,\{c\}}) \in
    \{ 0, 1, 2 \} \times \{ 0, 1, 2 \}
\end{align*}
and all the other parameters keep their inferred values.
%These parametrizations take the form of sets of $\param$ atoms.
% Note that $\{ 0, 2 \}$ belongs to the set of candidate parametrizations
% as no rule specifying that a parameter has to be an interval has been defined yet.
}%
\end{example}

>>>>

Cardinality rules turn out to be convenient to enumerate all possible parametrizations by creating multiple answer sets.
For functional purposes, a unique label is assigned to every possible set of resources of a given component.
Thus, we denote $\omega_p$ the set of resources of a given component $a$ labeled by $p$,
and naturally, $K_{a,\omega_p}$ is the related parameter.
%and in the following we note $K_{a,\omega_p}$ the parameter of component $a$ whose set of resources $\omega$ is assigned to the label $p$.
We note that labeling the sets of resources of a component is obviously equivalent to labeling its parameters.
Then, suppose that:
\begin{itemize}
  \item $\paramlabel(a, p)$ states that $p$ is a valid label for a set of resources of component $a$ (and therefore $K_{a,\omega_p}$ is a valid parameter);
  \item $\param(a, p, i)$ states that: $K_{a, \omega_p} = i$;
  \item $\inferedparam(a, p)$ states that the parameter inference of $K_{a, \omega_p}$ was conclusive (\pref{pps:param_K}).
\end{itemize}
It is thereby possible to enumerate the possible values of all parameters for which \pref{pps:param_K} was not conclusive, with the following cardinality rule:
\begin{align*}
  & 1\ \{\ \param(\var{A}, \var{P}, \var{I}) \mid \componentlevels(\var{A}, \var{I})\ \}\ \modMF{1}\ \la \\
  & \qquad\qquad\qquad \paramlabel(\var{A}, \var{P}), \nota \inferedparam(\var{A}, \var{P}).
\end{align*}
Indeed, this rule applies to any possible parameter $\var{P}$ of any component $\var{A}$ ($\paramlabel$) whose value is still unknown ($\nota \inferedparam$),
and states that any expression level $\var{I}$ of this component ($\componentlevels$)
is a candidate value for the parameter ($\param$).
\modMF{
Furthermore, the lower and upper bounds are both $1$,
which forces each enumerated parameter to have exactly one value.
}
%but no upper bound is specified ($\infty$) for the size of each parameter
%(which is in fact already bounded by the number of possible expression levels of the related component).
%each parameter contains at least $1$ value, as stated by the lower bound, and has no upper bound ($\infty$) but the number of expression levels of its component.
In other worlds, this cardinality rule creates as many answer sets as there are \emph{candidate} parametrizations
so that if $K_{a, \omega_p}$ could not be inferred by \pref{pps:param_K}, then
\modMF{
$K_{a, \omega_p} \in \segm{0}{l_a}$
}
(thus completely disregarding the notion of admissible parametrizations given in \pref{ssec:admissible-K}).

\begin{example}\label{ex:cardinality}
\modMF{%
In the scope of \pref{ex:infer-param-runningPH-1},
$K_{a,\{b\}}$ and $K_{a,\{c\}}$
could not be inferred by \pref{pps:param_K}.
The previous cardinality rule allows to produce 9 parametrizations,
in which these two parameters can take all possible values:
%(as, in this case, the assumptions of \pref{ssec:admissible-K} bring no new constraint):
\begin{align*}
  (K_{a,\{b\}}; K_{a,\{c\}}) \in
    \{ 0, 1, 2 \} \times \{ 0, 1, 2 \}
\end{align*}
and all the other parameters keep their inferred values.
%These parametrizations take the form of sets of $\param$ atoms.
% Note that $\{ 0, 2 \}$ belongs to the set of candidate parametrizations
% as no rule specifying that a parameter has to be an interval has been defined yet.
}%
\end{example}

### 2014-11-12 parts-discussion.tex
section 4.1
corrections
last minor corrections

We first propose an original inference of the Interaction Graph (IG) from a BRN
having its dynamics specified in the PH framework.
An IG gives a compact abstract representation of the influence of the components between each
others.
Then, based on a prior inference of Ren\'e Thomas' parametrization for BRNs from a PH model, we
delimit the set of admissible Thomas' parametrizations that are compatible with the PH dynamics,
and give arguments for their correctness.
A parametrization is compatible with the PH if its dynamics (in terms of possible transitions) is included in the PH dynamics.
The enumeration of such parametrizations is efficiently tackled using Answer Set Programming.
We illustrate the overall method with several results on large biological models.

Several extensions of the presented work are now to be considered.
First, the link between successively refined models of a system could be formally studied.
Indeed, it is convenient to refine a PH model by removing actions or adding cooperations;
the study and formalization of such process would allow to predict behavioral changes and lead to more accurate models.
Such work would also require further study on the inference of the IG in order to better understand and possibly avoid
the presence of some of the current spurious auto-edges.
Second, the inference of BRN multiplexes \cite{BernotMultiplexes} may be of practical interest 
as they allow to implicitly reduce the possible parametrizations by making cooperations appear
in the IG.
Because of its atomicity, the PH allows to specify a range of cooperations that cannot be
completely captured by a single instance of BRN multiplexes, then encouraging the inference of a set
of compatible ones.
Finally, in order to improve the performances in the IG inference, we will consider projection operations on
the PH structure to undo cooperations between components and reduce the cardinality of
configurations to explore by making the interactions independent.

\paragraph{Acknowledgement}
This work was partially supported by the Fondation Centrale Initiatives and
the French National Agency for Research (ANR-10-BLANC-0218 BioTempo project).
The work of LP was supported by the Swiss SystemsX.ch project.

>>>>

We first propose an original inference of the Interaction Graph (IG) from a BRN
having its dynamics specified in the PH framework.
An IG gives a compact abstract representation of the influence of the components between each
others.
Then, based on a prior inference of Ren\'e Thomas' parametrization for BRNs from a PH model, we
delimit the set of admissible Thomas' parametrizations that are compatible with the PH dynamics,
and give arguments for their correctness.
A parametrization is compatible with the PH if its dynamics (in terms of possible transitions) is included in the PH dynamics.
The enumeration of such parametrizations is efficiently tackled using Answer Set Programming.
We illustrate the overall method with several results
on \modMF{both small and} large biological models.

Several extensions of the presented work are now to be considered.
First, the link between successively refined models of a system could be formally studied.
Indeed, it is convenient to refine a PH model by removing actions or adding cooperations;
the study and formalization of such process would allow to predict behavioral changes and lead to more accurate models.
Second, the inference of BRN multiplexes \cite{BernotMultiplexes} may be of practical interest 
as they allow to implicitly reduce the possible parametrizations by making cooperations appear
in the IG.
Because of its atomicity, the PH allows to specify a range of cooperations that cannot be
completely captured by a single instance of BRN multiplexes, then encouraging the inference of a set
of compatible ones.
Finally, in order to improve the performances in the IG inference, we will consider projection operations on
the PH structure to undo cooperations between components and reduce the cardinality of
configurations to explore by making the interactions independent.

\paragraph{Acknowledgement}
This work was partially supported by the Fondation Centrale Initiatives and
the French National Agency for Research (ANR-10-BLANC-0218 BioTempo project).

### 2014-11-12 parts-example.tex
last minor corrections

The inference \modMF{and enumeration} methods described in this paper have been implemented as part of
\textsc{Pint}\footnote{Available at \url{http://loicpauleve.name/pint}}, which gathers PH related tools.
Our implementation mainly consists in ASP programs that are solved using
Clingo\footnote{Available at \url{http://potassco.sourceforge.net}}.
The IG and parameter inferences can be performed using the following command:\\
  \hspace*{\parindent}\texttt{ph2thomas -i model.ph -{}-dot ig.dot}\\
where \texttt{model.ph} is the PH model file in \textsc{Pint} format,
and \texttt{ig.dot} is an output file to write the inferred IG in DOT format.
The (possibly partial) inferred parametrization will be returned on the standard output.
\modMF{%
Instead of the \texttt{-{}-dot ig.dot} option, it is possible to specify
an alternative IG to perform the parameters inference on
(instead of using the inferred IG) with the \texttt{-{}-ig model.ig} option,
where \texttt{model.ig} is an IG model file in \textsc{Pint} format.
The admissible parametrizations enumeration is performed by adding the \texttt{-{}-enumerate}
option to the command.
%or \texttt{-{}-enumerate} instead to obtain only simple parameters (intervals of cardinality $1$).
}%

@@

The composition of all models and the results of the inferences are summarized in \pref{tb:computation}.
%We note that due to a very high number of parameters, no parameters inference could be performed on the $94$ and $104$ component models.
%These models would therefore be more efficiently studied as PH than as BRNs.
%Finally, we note that the complexity of the method is exponential in the number of regulators of one
%component and linear in the number of components.

\begin{table}[ht]
\begin{center}
  \begin{tabular}{r|c|c|c|c|c}
    \textbf{Model} & $\mathbf{|\Gamma|}$ & $\mathbf{|\PHs \setminus \Gamma|}$ & \textbf{$\Delta t$ IG} & \textbf{$\Delta t$ K} & $\mathbf{|K|}$
\\\hline\hline
    EGF receptor \cite{Sahin09} & $20$ & $22$ & <1s & <1s & 192
\\\hline
    T cell receptor \cite{Klamt06} & $40$ & $14$ & <1s & <1s & 143
\\\hline
    T cell receptor \cite{SaezRodriguez2007} & $94$ & $39$ & \modMF{100s} & \modMF{<1s} & \modMF{578}
\\\hline
    EGF receptor \cite{Samaga2009} & $104$ & $89$ & \modMF{200s} & \modMF{2.5s} & \modMF{27496}
  \end{tabular}
\end{center}
\caption{%
  Computation times and several pieces of information related to the IG and parametrization inferences of four biological models.
  The second column gives the number of components of each model and
  the third column gives the number of cooperative sorts used to model joint actions.
  The fourth (resp.~fifth) column gives the computation times of the IG inference (resp.~the parametrization inference).
  The last column gives the number of parameters in each model.
  }
\label{tb:computation}
\end{table}

>>>>

The inference \modMF{and enumeration} methods described in this paper have been implemented as part of
\textsc{Pint}\footnote{Available at \url{http://loicpauleve.name/pint}}, which gathers PH related tools.
Our implementation mainly consists in ASP programs that are solved using
Clingo\footnote{Available at \url{http://potassco.sourceforge.net}}.
The IG and parameter inferences can be performed using the following command:\\
  \hspace*{\parindent}\texttt{ph2thomas -i model.ph -{}-dot ig.dot}\\
where \texttt{model.ph} is the PH model file in \textsc{Pint} format,
and \texttt{ig.dot} is an output file to write the inferred IG in DOT format.
The (possibly partial) inferred parametrization will be returned on the standard output.
\modMF{%
Instead of the \texttt{-{}-dot ig.dot} option, it is possible to specify
an alternative IG to perform the parameters inference on
(instead of using the inferred IG) with the \texttt{-{}-ig model.ig} option,
where \texttt{model.ig} is an IG model file in \textsc{Pint} format.
The admissible parametrizations enumeration is performed by appending the \texttt{-{}-enumerate}
option to the command.
%or \texttt{-{}-enumerate} instead to obtain only simple parameters (intervals of cardinality $1$).
}%

@@

The composition of all models and the results of the inferences are summarized in \pref{tb:computation}.
\modMF{%
In every case, all parameters could be inferred and it was not necessary to
enumerate compatible parametrizations.
}%
%We note that due to a very high number of parameters, no parameters inference could be performed on the $94$ and $104$ component models.
%These models would therefore be more efficiently studied as PH than as BRNs.
%Finally, we note that the complexity of the method is exponential in the number of regulators of one
%component and linear in the number of components.

\begin{table}[ht]
\begin{center}
  \begin{tabular}{r|c|c|c|c|c}
    \textbf{Model} & $\mathbf{|\Gamma|}$ & $\mathbf{|\Delta|}$ & \textbf{IG inference} & \textbf{K inference} & $\mathbf{|K|}$
\\\hline\hline
    EGF receptor \cite{Sahin09} & $20$ & $22$ & <1s & <1s & 192
\\\hline
    T cell receptor \cite{Klamt06} & $40$ & $14$ & <1s & <1s & 143
\\\hline
    T cell receptor \cite{SaezRodriguez2007} & $94$ & $39$ & \modMF{100s} & \modMF{<1s} & \modMF{578}
\\\hline
    EGF receptor \cite{Samaga2009} & $104$ & $89$ & \modMF{200s} & \modMF{2.5s} & \modMF{27496}
  \end{tabular}
\end{center}
\caption{%
  Computation times and several pieces of information related to the IG and parametrization inferences of four biological models.
  The second column gives the number of components of each model and
  the third column gives the number of cooperative sorts used to model joint actions.
  The fourth (resp.~fifth) column gives the computation times of the IG inference (resp.~the parametrization inference).
  The last column gives the number of parameters in each model.
  }
\label{tb:computation}
\end{table}

### 2014-11-12 parts-grn.tex
corrections mineures
last minor corrections

\modMM{
The principle of qualitative modeling was introduced as synchronous Boolean networks by Stuart Kauffman on the one hand \cite{kauffman1969metabolic}, asynchronous Ren\'e Thomas’ networks on the other \cite{Thomas73}. Both models have been considered of interest and led to numerous works. In the following, we choose to focus on  Thomas' modelling, with its asynchronous, unitary and multi-level semantics.}

\modMM{In this paragraph, we} concisely present Thomas' modeling of a Biological Regulatory Network (BRN) and its dynamics, merely inspired by
\cite{thomas1990biological, Richard06,BernotSemBRN}.
%In order to enlarge the class of Thomas' models compatible with PH dynamics (w.r.t.~the presented
%inference),
%we propose the notion of unsigned edge modeling an interaction whose nature (activation or inhibition) is undetermined,
%and we extend the classical parametrization formalism by setting parameters to intervals of values instead of single values.
%We briefly discuss these additions.

@@

\modMM{The IG is not sufficient to unambiguously define the dynamics of the system. In particular, a component may be regulated by both a positive and negative interaction derived from two different components. To refine the dynamics, a parametrization had been added to the model \cite{snoussi1989qualitative}. This parametrization allows to define focal points, \ie it clarifies the level towards which a component evolves depending on the expression of its regulators.
Some of the information included in the IG and the parametrization may appear as slighltly redundant. Especially, if the single parametrization were enriched with additional constraints (\eg monotonicity requiring an ordering on the value of parameters), then it may directly capture the signs. However, distinguishing the IG and parametrization allows to decouple the graph (which is generally one of the prior knowledge from the biologists) from the constraints on its dynamics. It opened the way to numerous studies (\eg necessary condition on sustained oscillations) solely based on the structure of the IG (\eg the analysis of positive or negative circuits).}

@@

\begin{remark}[Parameters equivalence]
\label{rem:K-equiv}
\modLP{%
In the scope of the asychronous dynamics (\pref{def:dynamics}), one can remark
that some parameters values can be equivalent, \ie the resulting dynamics is
the same.
It is notably the case with auto-regulations:
if $\GRNedge{a}{s}{t}{a}$,
then for any $\omega\subseteq \GRNreg{a}$,
two values $v$ and $v'$ of the parameter 
$K_{a,\omega\setminus \{a\}}$ are equivalent
if $v\geq t$ and $v'\geq t$.
Similary,
two values $v$ and $v'$ of the parameter
$K_{a,\omega\cup\{a\}}$ are equivalent 
if $v<t$ and $v'<t$.}
\end{remark}

\paragraph{Interaction Graph of the dynamics}
From the dynamics specification of a BRN, one can infer back the IG that contain only the functional
regulations.
Let us define
$f^a(s) = s[a]$ if $s[a] = K_{a, \GRNres{a}{s}}$
and
$f^a(s) = s[a]+\delta^a(s)$ otherwise.
A positive (resp. negative) interaction from $b$ to $a$ is inferred if there exists a state $s$ such
that increasing $b$ in $s$ would increase (resp. decrease) $f^a$; in other words, if
there exists a state $s$ so that:
$f^a(s\{b_i\}) < \text{(resp. $>$) } f^a(s\{b_{i+1}\})$, 
where $i < l_b$ and $s\{b_i\}$ denotes the state $s$ where the component $b$ is assigned to $b_i$.
\modLP{If all the interactions from $b$ to $a$ are positive (resp. negative), there is a positive
(resp. negative) edge from $b$ to $a$ in the IG, denoting a monotonous influence of $b$ on $a$.
In the case when there exists two differently signed interactions, the edge in the IG is said
non-monotonous.
This abstract representation of the dynamical system can then} be used to derive global properties on
the dynamics (\eg \cite{Richard2010378,PR11-SASB}).

>>>>

\modMM{
The principle of qualitative modeling was introduced as synchronous Boolean networks by Stuart Kauffman on the one hand \cite{kauffman1969metabolic}, and asynchronous Ren\'e Thomas' networks on the other \cite{Thomas73}. Both models have been considered of interest and led to numerous works. In the following, we choose to focus on  Thomas' modelling, with its asynchronous, unitary and multi-level semantics.}

\modMM{In this section, we} concisely present Thomas' modeling of a Biological Regulatory Network (BRN) and its dynamics, merely inspired by
\cite{thomas1990biological, Richard06,BernotSemBRN}.
%In order to enlarge the class of Thomas' models compatible with PH dynamics (w.r.t.~the presented
%inference),
%we propose the notion of unsigned edge modeling an interaction whose nature (activation or inhibition) is undetermined,
%and we extend the classical parametrization formalism by setting parameters to intervals of values instead of single values.
%We briefly discuss these additions.

@@

\modMM{The IG is not sufficient to unambiguously define the dynamics of the system. In particular, a component may be regulated by both a positive and negative interaction derived from two different components. To refine the dynamics, a parametrization has been added to the model \cite{snoussi1989qualitative}. This parametrization allows to define focal points, \ie it clarifies the level towards which a component evolves depending on the expression of its regulators.
Some of the information included in the IG and the parametrization may appear as slightly redundant. Especially, if the single parametrization were enriched with additional constraints (\eg monotonicity requiring an ordering on the value of parameters), then it may directly capture the signs. However, distinguishing the IG and parametrization allows to decouple the graph (which is generally one of the prior knowledge from the biologists) from the constraints on its dynamics. It opened the way to numerous studies (\eg necessary condition on sustained oscillations) solely based on the structure of the IG (\eg the analysis of positive or negative circuits).}

@@

\begin{remark}[Parameters equivalence]
\label{rem:K-equiv}
\modLP{%
In the scope of the asychronous dynamics (\pref{def:dynamics}), one can remark
that some parameters values can be equivalent, \ie the resulting dynamics is
the same.
It is notably the case with self-regulations:
if $\GRNedge{a}{s}{t}{a}$,
then for any $\omega\subseteq \GRNreg{a}$,
two values $v$ and $v'$ of the parameter 
$K_{a,\omega\setminus \{a\}}$ are equivalent
if $v\geq t$ and $v'\geq t$.
Similary,
two values $v$ and $v'$ of the parameter
$K_{a,\omega\cup\{a\}}$ are equivalent 
if $v<t$ and $v'<t$.}
\end{remark}

\paragraph{Interaction Graph of the dynamics}
From the dynamics specification of a BRN, one can infer back the IG that contain only the functional
regulations.
Let us define
$f^a(s) = s[a]$ if $s[a] = K_{a, \GRNres{a}{s}}$
and
$f^a(s) = s[a]+\delta^a(s)$ otherwise.
A positive (resp. negative) interaction from $b$ to $a$ is inferred if there exists a state $s$ such
that increasing $b$ in $s$ would increase (resp. decrease) $f^a$; in other words, if
there exists a state $s$ so that:
$f^a(s\{b_i\}) < \text{(resp. $>$) } f^a(s\{b_{i+1}\})$, 
where $i < l_b$ and $s\{b_i\}$ denotes the state $s$ where the component $b$ is assigned to $i$.
\modLP{If all the interactions from $b$ to $a$ are positive (resp. negative), there is a positive
(resp. negative) edge from $b$ to $a$ in the IG, denoting a monotonous influence of $b$ on $a$.
In the case when there exists two differently signed interactions, the edge in the IG is said
non-monotonous.
This abstract representation of the dynamical system can then} be used to derive global properties on
the dynamics (\eg \cite{Richard2010378,PR11-SASB}).
